%&latex -*- latex -*-

\title{Scsh Reference Manual}
\subtitle{For scsh release 0.6.6}
\author{Olin Shivers, Brian D.~Carlstrom, Martin Gasbichler, and Mike Sperber}
\date{March 2004}

\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Some code-changes for tex2page and latex output. NF
\texonly
\chapter*{Acknowledgements}
\endtexonly
\htmlonly
\\ \ex{Acknowledgements} \\ \\
\endhtmlonly

Who should I thank?
My so-called ``colleagues,'' who laugh at me behind my back,
    all the while becoming famous on {\em my\/} work?
My worthless graduate students, whose computer skills appear to
    be limited to downloading bitmaps off of netnews?
My parents, who are still waiting for me to quit ``fooling around with
    computers,'' go to med school, and become a radiologist?
My department chairman, a manager who gives one new insight into
    and sympathy for disgruntled postal workers?

My God, no one could blame me---no one!---if I went off the edge and just
    lost it completely one day.
I couldn't get through the day as it is without the Prozac and Jack Daniels
    I keep on the shelf, behind my Tops-20 JSYS manuals.
    I start getting the shakes real bad around 10am, right before my
    advisor meetings. A 10 oz.\ Jack 'n Zac helps me get through the
    meetings without one of my students winding up with his severed head
    in a bowling-ball bag. They look at me funny; they think I twitch a
    lot. I'm not twitching. I'm controlling my impulse to snag my 9mm
    Sig-Sauer out from my day-pack and make a few strong points about
    the quality of undergraduate education in Amerika.

If I thought anyone cared, if I thought anyone would even be reading this,
I'd probably make an effort to keep up appearances until the last possible
moment. But no one does, and no one will. So I can pretty much say exactly
what I think.

Oh, yes, the {\em acknowledgements.}
I think not. I did it. I did it all, by myself.
\begin{flushright}
\begin{tabular}{l}
Olin Shivers    \\
Cambridge       \\
September 4, 1994
\end{tabular}
\end{flushright}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\tableofcontents
